---
title: "Lab 5"
format: html
editor: visual
embed-resources: true
fig-height: 10
---

## Set Up

Load the met data and also the station data, then merge the data as we did during the lecture.

```{r}
# Download the data
stations <- read.csv("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations$USAF <- as.integer(stations$USAF)

# Dealing with blanks and 999999
stations$USAF[stations$USAF == 999999] <- NA
stations$CTRY[stations$CTRY == ""] <- NA
stations$STATE[stations$STATE == ""] <- NA

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, c('USAF', 'CTRY', 'STATE')])

# Dropping NAs
stations <- stations[!is.na(stations$USAF), ]

# Removing duplicates
stations <- stations[!duplicated(stations$USAF), ]

#Load Libraries 
library(data.table)
library(dplyr)
library(dtplyr)
library(leaflet)
library(ggplot2)

# Load met_all Data
met_all <- fread("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz")
met_all <- met_all %>% rename(USAF = USAFID)
met <- met_all %>% left_join(stations, by = "USAF")

# Remove NAs & Duplicates
met <- met[!is.na(met$USAF), ]
met <- met[!duplicated(met$USAF), ]

```

## Question 1

What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the quantile() function. 

Do these three coincide?
    - The three representative stations did not coincide, because temperature, wind speed, and atmospheric pressure vary independently across geography and time.
    
```{r}
# Compute median values across all US observations
med_temp  <- quantile(met$temp, probs = 0.5, na.rm = TRUE)
med_wind  <- quantile(met$wind.sp, probs = 0.5, na.rm = TRUE)
med_press <- quantile(met$atm.press, probs = 0.5, na.rm = TRUE)

med_temp
med_wind
med_press


# Find the station closest to each median
rep_temp_station <- met %>%
  mutate(diff = abs(temp - med_temp)) %>%
  slice(which.min(diff)) %>%
  pull(USAF)

rep_wind_station <- met %>%
  mutate(diff = abs(wind.sp - med_wind)) %>%
  slice(which.min(diff)) %>%
  pull(USAF)

rep_press_station <- met %>%
  mutate(diff = abs(atm.press - med_press)) %>%
  slice(which.min(diff)) %>%
  pull(USAF)

# Display the three representative stations
c(
  temp_station  = rep_temp_station,
  wind_station  = rep_wind_station,
  press_station = rep_press_station
)


```

## Question 2

Calculate the median latitude and longitude for stations in each state.

```{r}
# Compute median lat/lon for each state
state_medians <- met %>%
  group_by(STATE) %>%
  summarize(
    med_lat = median(lat, na.rm = TRUE),
    med_lon = median(lon, na.rm = TRUE)
  )

head(state_medians)

```

## Question 3 

For each state, identify the station that is closest to the mean latitude and longitude for stations in that state. Combining these with the coordinates you identified in the previous question, use leaflet() to visualize all ~100 points in the same figure, applying different colors for those identified in this question.

```{r}
# group the mean lat/lon per state
state_means <- met %>%
  group_by(STATE) %>%
  summarize(
    mean_lat = mean(lat, na.rm = TRUE),
    mean_lon = mean(lon, na.rm = TRUE)
  )

# Identify the station closest to each mean coordinate
closest_stations <- met %>%
  inner_join(state_means, by = "STATE") %>%
  mutate(dist = sqrt((lat - mean_lat)^2 + (lon - mean_lon)^2)) %>%
  group_by(STATE) %>%
  slice(which.min(dist)) %>%
  ungroup()

# Visualize both data sets with leaflet
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    data = state_medians,
    lng = ~med_lon, lat = ~med_lat,
    color = "blue", radius = 5,
    popup = ~paste("Median:", STATE)
  ) %>%
  addCircleMarkers(
    data = closest_stations,
    lng = ~lon, lat = ~lat,
    color = "red", radius = 5,
    popup = ~paste("Closest to Mean:", STATE)
  )

```

## Question 4

Compute each state’s average temperature and use that to classify them according to the following criteria:

low: temp < 20
Mid: temp >= 20 and temp < 25
High: temp >= 25
Now generate a large summary table with states/observations split into the three categories listed above. For each group, compute the following:

Number of states
Number of entries (records)
Number of NA entries
Number of stations
Mean temperature
Mean wind-speed
Mean atmospheric pressure

```{r}
# Compute each state’s average temperature
state_temp <- met %>%
  group_by(STATE) %>%
  summarize(
    mean_temp = mean(temp, na.rm = TRUE),
    mean_wind = mean(wind.sp, na.rm = TRUE),
    mean_press = mean(atm.press, na.rm = TRUE),
    n_records = n(),
    n_na = sum(is.na(temp)),
    n_stations = n_distinct(USAF)
  )

# Classify states into Low / Mid / High temperature groups
state_temp <- state_temp %>%
  mutate(
    category = case_when(
      mean_temp < 20 ~ "Low",
      mean_temp >= 20 & mean_temp < 25 ~ "Mid",
      mean_temp >= 25 ~ "High"
    )
  )

# Build the large summary table
summary_table <- state_temp %>%
  group_by(category) %>%
  summarize(
    n_states = n(),
    total_records = sum(n_records),
    total_na = sum(n_na),
    total_stations = sum(n_stations),
    mean_temp = mean(mean_temp),
    mean_wind = mean(mean_wind),
    mean_press = mean(mean_press)
  )
head(summary_table)
```