---
title: "Lab 3"
format: html
editor: visual
embed-resources: true
fig-height: 10
---

## File Upload

```{r}
met_all <- read.csv("~/PM 556/PM556-/met_all.gz")
```

## Question 2

Check the dimensions, headers, footers. How many columns, rows are there?

```{r}
dim(met_all)
head(met_all)
tail(met_all)
```

## Question 3

Take a look at the variables.

```{r}
str(met_all)
```

## Question 4

Take a closer look at the key variables.

```{r}
table(met_all$year)
table(met_all$day)
table(met_all$hour)
summary(met_all$temp)
summary(met_all$elev)
summary(met_all$wind.sp)

# Check the data for potential missing values 

met_all$elev[met_all$elev==9999.0] <- NA
summary(met_all$elev)

# Remove observations that contain the minimum temperature of -40C. 

met_all <- met_all[met_all$temp > -40, ]
met2 <- met_all[order(met_all$temp), ]
head(met2)
```

## Question 5

Check the data against an external data source. Fix any problems that arise in your checks.

```{r}
met_all <- met_all[met_all$temp > -15, ]
met2 <- met_all[order(met_all$temp), ]
head(met2)

# Summarize that we removed temperatures colder than -15C. The new dataset has a minimum temp of -3C, which is reasonable.
```

## Question 6

Calculate summary statistics. Select the weather station with maximum elevation; look at the correlation between temperature and wind speed; look at the correlation between temperature and wind speed with hour and day of the month.

```{r}
elev <- met_all[met_all$elev == max(met_all$elev, na.rm = TRUE), ]
summary(elev)


cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use="complete")
cor(elev$wind.sp, elev$day, use="complete")
cor(elev$wind.sp, elev$hour, use="complete")
cor(elev$temp, elev$day, use="complete")
```

## Question 7

Exploratory graphs.

```{r}
hist(met_all$elev, breaks=100)
hist(met_all$temp)
hist(met_all$wind.sp)


# Install Package and run pipe operator 

library(leaflet)
library(magrittr)

# Look at where the weather station with highest elevation is located

leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)

# Look at the time series of temperature and wind speed at this location. For this we will need to create a date-time variable for the x-axis.

library(lubridate)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
elev <- elev[order(elev$date), ]
head(elev)

# Plot the time series of temperature and wind speed.

plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```

## Question 8

I don't necessarily ave questions regarding the data, however here are two that seemed interesting:

1.  Why do some temperature readings show extreme values like `-40°C` or `-17.2°C`?
2.  Were all stations using the same hardware and calibration standards?
